Detecting Fair Queuing for Better Congestion Control

### 2021-02-11 paper/paper_ieee.tex
Add paper in IEEE format

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage[english]{babel}
\usepackage{xcolor}
\usepackage[nomain, toc, acronym]{glossaries}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
\renewcommand\UrlFont{\color{blue}\rmfamily}

@@

\usepackage{amsmath,amssymb,amsfonts}
\newcommand{\mynote}[3]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\small$\blacktriangleright$\textsf{\emph{\color{#3}{#2}}}$\blacktriangleleft$}}

\newcommand{\todo}[1]{\mynote{TODO}{#1}{red}}
\newcommand{\noteMax}[1]{\mynote{Max}{#1}{green}}

@@

\begin{document}
%
%JF: \title{Detecting Fair Queuing to Enable Better Congestion Control}
%JF: Alternative Vorschläge für Titel, etwas flüssiger:
\title{Optimizing Congestion Control Through Fair Queuing Detection}
%JF: \title{Optimum Congestion Control Selection Through Fair Queuing Detection}

%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\ifx\anonymize\tempYes
\author{Anonymous authors}
%
\authorrunning{Anonymous authors}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\else
\author{Maximilian Bachl\inst{1}\orcidID{0000-0001-9482-7428} \and
Joachim Fabini\inst{2,3}\orcidID{0000-0002-8285-1591} \and
Tanja Zseby\inst{3}\orcidID{0000-0002-5391-467X}}
%
\authorrunning{M. Bachl et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{TU Wien, Vienna, Austria; 
\email{firstname.lastname@tuwien.ac.at}}
\fi

\maketitle              % typeset the header of the contribution

%JF: FIXME: A fundamental problem that we should perhaps address in a brief statement: The proposed method does not work for cases in which load balancers are on the path. Distinct flows are likely to result in distinct hashes/flowIDs, meaning that the balancer will potentially assign them to distinct paths. So we may add a restrictive statement to the discussion that we assume that all flows share the same bottleneck? In addition we may state that no service/port priorization is in place.
%JF Any other idea?

\begin{abstract}
Low delay is an explicit requirement for applications such as cloud gaming and video conferencing. Delay-based congestion control can achieve the same throughput but significantly smaller delay than loss-based one and is thus ideal for these applications. However, when a delay- and a loss-based flow compete for a bottleneck, the loss-based one can monopolize all the bandwidth and starve the delay-based one. Fair queuing at the bottleneck link solves this problem by assigning an equal share of the available bandwidth to each flow. However, so far no end host based algorithm to detect fair queuing exists. Our contribution is the development of an algorithm that detects fair queuing at flow startup and chooses delay-based congestion control if there is fair queuing. Otherwise, loss-based congestion control can be used as a backup option. Results show that our algorithm reliably detects fair queuing and can achieve low delay and high throughput in case fair queuing is detected.

\keywords{Congestion Control  \and Active Measurements \and Fair Queuing \and Queue Management.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:introduction}

@@

\subsubsection{\gls{cc} using Active Measurements}
Several approaches have been recently proposed which aim to incorporate active measurements into \glspl{cca}. \cite{dong_pcc_2015,dong_pcc_2018} introduced the \gls{cca} \textit{PCC}, which uses active measurements to find the sending rate which optimizes a given reward function. \cite{cardwell_bbr:_2016} introduce the \textit{BBR} \gls{cca}, which uses active measurement to create a model of the network connection. This model is then used to determine the optimal sending rate. \cite{bachl_rax_2019} use Reinforcement Learning to learn \gls{cc} policies. This is done by performing active measurements by varying the congestion window to evaluate which value gives the highest throughput and lowest packet loss. Another approach proposed by \cite{goyal_elasticity_2018} aims to differentiate bulk transfer flows, which they deem ``elastic'' because they try to grab as much bandwidth as available, from other flows that send at a constant rate, which they call ``unelastic''. They use this elasticity detection to determine if delay-based \gls{cc} can be used or not. If cross traffic is deemed elastic then they use classic loss-based \gls{cc} and otherwise delay-based \gls{cc}.

\subsubsection{Flow startup techniques} Besides the classical TCP slow start algorithm \cite{stevens_tcp_1997} some other proposals were made such as Quick Start \cite{jain_quick-start_2019}, which aims to speed up flow start by routers informing the end point, which rate they deem appropriate. \cite{jarvinen_congestion_2019} investigate the impact of \gls{aqm} and \gls{cc} during flow startup. They inspect a variety of different \gls{aqm} algorithms and also investigate \gls{fq}. \cite{kuhlewind_chirping_2010} propose ``chirping'' for flow startup to estimate the available bandwidth. This works by sending packets with different gaps between them to verify what the actual link speed is. \cite{mittal_recursively_2014} propose speeding up flow startup by using packets of different priority. The link is flooded with low-priority packets and in case the link is not fully utilized during startup, these low-priority packets are going to be transmitted, allowing the starting flow to achieve high throughput. If the link is already saturated, the low-priority packets are going to be discarded and no other flow suffers.

%JF: Evtl.statt subsubsection Absätze mit Titel fett verwenden um doch einigen Platz zu sparen?
\subsubsection{Detecting \gls{aqm} mechanisms} There are few publications that investigate methods to detect the presence of an \gls{aqm}. \cite{kargar_bideh_tada_2016} propose a method to detect and distinguish certain popular \gls{aqm} mechanisms. However, unlike our work, it doesn't consider \gls{fq}. \cite{baykal_detection_2017} propose a machine learning based approach to fingerprint \gls{aqm} algorithm. \gls{fq} is not considered, too.

\section{Concept}
\label{sec:concept}

@@

%\begin{figure}
%\centering
%\includegraphics[width=1\columnwidth]{{"figures/fq_illustration_throughput"}.pdf}
%\caption{Sending rate}
%\label{fig:throughput}
%\end{figure}
%
%\begin{figure}
%\includegraphics[width=1\columnwidth]{{"figures/fq_illustration_goodput_no_fq"}.pdf}
%\caption{Receiving rate (goodput) in case the bottleneck link doesn't use \gls{fq}. Flow 2, which sends more also manages to get more data through the bottleneck.}
%\label{fig:goodput_no_fq}
%\end{figure}
%
%\begin{figure}
%\includegraphics[width=1\columnwidth]{{"figures/fq_illustration_goodput_fq"}.pdf}
%\caption{Receiving rate (goodput) in case the bottleneck link uses \gls{fq}. Flow 2, which sends more gets as much data through the bottleneck as flow 1.}
%\label{fig:goodput_fq}
%\end{figure}

\begin{figure}
\centering
\subfloat[Sending rate]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_throughput"}.pdf}
\label{fig:throughput}}
\subfloat[Receiving rate (no \gls{fq}).]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_goodput_no_fq"}.pdf}
\label{fig:goodput_no_fq}}
\subfloat[Receiving rate (\gls{fq}).]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_goodput_fq"}.pdf}
\label{fig:goodput_fq}}
\caption{An example illustrating our proposed flow startup mechanism. Figure \ref{fig:throughput} shows the sending rate, \ref{fig:goodput_no_fq} the receiving rate in case there's no \gls{fq} and \ref{fig:goodput_fq} the receiving rate if there is \gls{fq}.}
\label{fig:illustration}
\end{figure}

The algorithm to determine the presence of \gls{fq} works as follows (seen from the sender's point of view):
\begin{enumerate}
\item Launch two concurrent flows, flow 1 and flow 2. Initialize flow 1 with some starting sending rate and flow 2 with two times that rate. 
\item After every \gls{rtt}, if no packet loss occurred, double the sending rate of both flows. 
\item If packet loss occurred in both flows in the previous \gls{rtt}, calculate the following metric: 
\begin{align}
\textit{loss ratio} &= \frac{\left(\frac{\textit{receiving rate of flow 1}}{\textit{sending rate of flow 1}}\right)}{\left(\frac{\textit{receiving rate of flow 2}}{\textit{sending rate of flow 2}}\right)}\end{align}

%JF: Perhaps abbreviate: sendRate flow1, recvRate flow1...
This loss ratio metric indicates if flow 2, which has a higher sending rate, achieves a higher receiving rate (goodput) than flow 1. If this is the case, there's no \gls{fq}, otherwise there is.

It is necessary to wait for packet loss in both connections because only then it is certain that the link is saturated. Our algorithm assumes that the link is saturated. 
\item If \gls{fq} is used, the $\textit{loss ratio}$ is 2, otherwise it is 1. We choose 1.5 as the cutoff value. 
\item Since the presence or absence of \gls{fq} is now known the appropriate \gls{cca} can be launched. 
\end{enumerate}

Figure~\ref{fig:throughput}, \ref{fig:goodput_no_fq} and \ref{fig:goodput_fq} schematically illustrate this mechanism. We assume a sender being connected to a receiver, with a bottleneck link between them. On this bottleneck link, the queue is either managed by \gls{fq} or by a shared queue (no \gls{fq}). Figure~\ref{fig:throughput} shows how the sending rate (observed before the bottleneck link) is doubled after each \gls{rtt}. Furthermore, it shows that flow 2 has double the sending rate of flow 1. %\noteMax{Figures take up too much space. Is it better to put them next to each other? The actual problem is the nonsensical Springer format.}.

@@

If \gls{fq} is detected, a simple delay-based \gls{cc} is used from then on. Our algorithm is based on the two following simple rules:
\begin{enumerate}
\item If $\textit{current rtt} \leq \textit{smallest rtt ever measured on the connection} + 5\textit{ms}$ then increase the sending rate by 1\%.
\item Otherwise ($\textit{current rtt} > \textit{smallest rtt ever measured on the connection} + 5\textit{ms}$) decrease the sending rate by 5\%.
\end{enumerate}

This \gls{cca} is meant as a simple proof-of-concept to demonstrate the efficacy of our \gls{fq} detection mechanism but we do not claim it to be optimal in any way.

@@

If the absence of \gls{fq} is detected, a loss based \gls{cc} is used as a fallback. We use the algorithm \textit{PCC} proposed by \cite{dong_pcc_2015}. In simplified terms, this \gls{cca} aims to maximize a utility function, which strives for high throughput and low packet loss. However, it is reasonably aggressive and does not starve when competing with other loss-based \glspl{cca} such as Cubic \cite{ha_cubic_2008}.

@@

We base our implementation on the code of \textit{PCC Vivace} \cite{dong_pcc_2018}, which in turn is an improved version of PCC. The PCC code is based on the \textit{UDT} library \cite{gu_udt_2007}, which is a library that provides a reliable transport protocol similar to TCP on top of UDP. This approach is similar to QUIC \cite{iyengar_quic_2018}. An advantage of implementing our approach on top of UDP as opposed to TCP is that it is easier to develop and modify since no kernel module needs to be created and that it is more secure since the code is isolated in a user space process.

@@

Our approach of detecting \gls{fq} relies on using two concurrent flows between the sender and the receiver. During flow startup, user data is transmitted in two flows that the receiver must detect and reassemble. This may be a challenge for deployments since solutions that require cooperation of several instances are more difficult to deploy at a wide scale in the Internet. %This can be seen for example when looking at the deployment of IPv6, which requires the participation of every router on the way and every end device. 
%JF: FIXME: I'm sceptical that the examples below work well in all scenarios. Distinct flows end up on distinct paths, so they do not necessarily use the same bottleneck.
However, multipath TCP \cite{handley_tcp_2012} and QUIC, which enable recombining data of several flows, are becoming increasingly prevalent and Google's QUIC implementation has multipath support \cite{google_quiche_2020}. We thus argue that the actual obstacles to deployment of our solution are constantly decreasing since QUIC (mostly Google traffic) already accounts for around %$\frac{1}{10}$ 
$10\%$ of all Internet traffic \cite{ruth_first_2018} and is growing. Moreover, Apple's macOS and iOS support multipath TCP \cite{apple_about_2018} support.

@@

For the evaluation we use the network emulation library \texttt{py-virtnet}\footnote{\url{https://pypi.org/project/py-virtnet/}}. For the experiments with \gls{fq} we use the \texttt{fq} module for Linux. This module keeps one drop tail queue for each flow with a configurable maximum size of packets. However, the design of our \gls{fq} detection algorithm does not assume any particular queue management algorithm being used for the queue of each flow and thus is also compatible with more sophisticated \gls{aqm} mechanisms which were explicitly developed to be used in conjunction with \gls{fq} \cite{taht_flow_2018,hoiland-jorgensen_piece_2018,bachl_cocoa_2019,bachl_lfq_2020}. For the experiments with a shared queue we use \texttt{pfifo}, which is a simple drop tail queue which can contain up to a fixed number of packets.

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_just_one_flow_vegas_udp_port9000_fq_10_10_20_1600183241479.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_just_one_flow_vegas_udp_port9000_fq_10_10_20_1600183241479.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by our delay-based \gls{cca} on a link with a speed of 10\,Mbps, a delay of 10\,ms and a buffer size of 100 packets.}
\label{fig:just_one_flow_udp}
\end{figure}

First, we demonstrate empirically that delay-based \gls{cc} actually has a benefit over loss-based one. On one hand, \autoref{fig:just_one_flow_tcp} shows that the loss-based \gls{cc} achieves a throughput of almost 10\,Mbps on a 10\,Mbps link, which means full utilization and an \gls{rtt} that is  higher than 100\,ms. On the other hand, \autoref{fig:just_one_flow_udp} shows an example flow using our simple delay-based \gls{cca} and a slow start algorithm similar to the one used by the classic TCP \gls{cca} \textit{New Reno} \cite{allman_tcp_1999}. While the delay-based flow also achieves very high utilization, delay rarely exceeds 20\,ms, meaning that it is more than 5 times lower than the one of the loss-based \gls{cca} of \autoref{fig:just_one_flow_tcp}.

\subsection{Comparing flows with different Congestion Controls and AQM mechanisms}
In this scenario, we show an example of a flow with our mechanism that shares a bottleneck link with a flow using a loss-based congestion control. 
%JF: More compelling would be the use case when the delay-based flow is started first and the late-comer loss-based preempts it.

\subsubsection{No \gls{fq}}
The examples in \autoref{fig:pfifo_tcp} and in \autoref{fig:pfifo_udp} show the loss-based \gls{cca} Cubic compete against our delay-based \gls{cca} on a link without \gls{fq}. It is important to note that we do not use our \gls{fq} detection mechanism here but instead choose delay-based \gls{cc} from the beginning to demonstrate the starvation of delay-based flows that happens when they compete against loss-based ones without \gls{fq}. In the scenario, the loss-based flow starts 5 seconds earlier and after 5 seconds, the delay-based flow joins. While the delay-based flow gains some share of the link during startup, it is pushed away by the loss-based flow later on and ``starves''. We start the loss-based flow before to make sure that it has sufficient time to fill up the link.

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_competing_flow_pfifo_vegas_udp_port9000_pfifo_10_50_20_1600182743408.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_competing_flow_pfifo_vegas_udp_port9000_pfifo_10_50_20_1600182743408.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by our delay-based \gls{cca} on a link with a speed of 50\,Mbps, a delay of 10\,ms and a buffer size of 100 packets. The bottleneck is controlled by a shared queue (no \gls{fq}) and is shared with the flow of \autoref{fig:pfifo_tcp}.}
\label{fig:pfifo_udp}
\end{figure}

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_competing_flow_fq_udp_port9000_fq_10_50_20_1600182657352.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_competing_flow_fq_udp_port9000_fq_10_50_20_1600182657352.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by our delay-based \gls{cca} on a link with a speed of 50\,Mbps, a delay of 10\,ms and a buffer size of 100 packets. The bottleneck is controlled by \gls{fq} and is shared with the flow of \autoref{fig:fq_tcp}.}
\label{fig:fq_udp}
\end{figure}

>>>>

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage[inline]{enumitem}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage[pdfauthor={Maximilian Bachl, Joachim Fabini, Tanja Zseby}]{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{subfig}
\usepackage{tikz}
\usepackage{float}

\usepackage[nomain, toc, acronym]{glossaries}
\glsdisablehyper

\newcommand\note[2]{{\color{#1}#2}}
\newcommand\todo[1]{{\note{red}{TODO: #1}}}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Detecting Fair Queuing for Better Congestion Control}

\author{\IEEEauthorblockN{Maximilian Bachl, Joachim Fabini, Tanja Zseby}
\IEEEauthorblockA{Technische Universität Wien\\
firstname.lastname@tuwien.ac.at}}

%\date{\today}

% \IEEEoverridecommandlockouts
% \IEEEpubid{\begin{minipage}[t]{\textwidth}\ \\[10pt]
%         \centering\normalsize{xxx-x-xxxx-xxxx-x/xx/\$31.00 \copyright 2018 IEEE}
% \end{minipage}}

% \renewcommand*{\bibfont}{\footnotesize}

%\newcommand\copyrighttext{%
%  \footnotesize \textcopyright 2020 IEEE. Personal use of this material is permitted.
%  Permission from IEEE must be obtained for all other uses, in any current or future
%  media, including reprinting/republishing this material for advertising or promotional
%  purposes, creating new collective works, for resale or redistribution to servers or
%  lists, or reuse of any copyrighted component of this work in other works.}
%\newcommand\copyrightnotice{%
%\begin{tikzpicture}[remember picture,overlay]
%\node[anchor=south,yshift=10pt] at (current page.south) {\fbox{\parbox{\dimexpr\textwidth-\fboxsep-\fboxrule\relax}{\copyrighttext}}};
%\end{tikzpicture}%
%}

\maketitle%
%\copyrightnotice

 \thispagestyle{plain}
 \pagestyle{plain}

@@

\begin{abstract}
Low delay is an explicit requirement for applications such as cloud gaming and video conferencing. Delay-based congestion control can achieve the same throughput but significantly smaller delay than loss-based one and is thus ideal for these applications. However, when a delay- and a loss-based flow compete for a bottleneck, the loss-based one can monopolize all the bandwidth and starve the delay-based one. Fair queuing at the bottleneck link solves this problem by assigning an equal share of the available bandwidth to each flow. However, so far no end host based algorithm to detect fair queuing exists. Our contribution is the development of an algorithm that detects fair queuing at flow startup and chooses delay-based congestion control if there is fair queuing. Otherwise, loss-based congestion control can be used as a backup option. Results show that our algorithm reliably detects fair queuing and can achieve low delay and high throughput in case fair queuing is detected. 
\end{abstract}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
%\begin{abstract}
%  A clear and well-documented \LaTeX\ document is presented as an
%  article formatted for publication by ACM in a conference proceedings
%  or journal publication. Based on the ``acmart'' document class, this
%  article presents and explains many of the common variations, as well
%  as many of the formatting elements an author may use in the
%  preparation of the documentation of their work.
%\end{abstract}

\maketitle

\glsresetall
\newacronym{cc}{CC}{Congestion Control}
\newacronym{cca}{CCA}{Congestion Control Algorithm}
\newacronym{aqm}{AQM}{Active Queue Management}
\newacronym{rtt}{RTT}{Round Trip Time}
\newacronym{fq}{FQ}{Fair Queuing}
\newacronym{bdp}{BDP}{Bandwidth Delay Product}
\newacronym{qoe}{QoE}{Quality of Experience}

@@

\section{Introduction}
\label{sec:introduction_fq}

@@

\subsubsection{\gls{cc} using Active Measurements}
Several approaches have been recently proposed which aim to incorporate active measurements into \glspl{cca}. \citep{dong_pcc_2015,dong_pcc_2018} introduced the \gls{cca} \textit{PCC}, which uses active measurements to find the sending rate which optimizes a given reward function. \citep{cardwell_bbr:_2016} introduce the \textit{BBR} \gls{cca}, which uses active measurement to create a model of the network connection. This model is then used to determine the optimal sending rate. An approach proposed by \citep{goyal_elasticity_2018} aims to differentiate bulk transfer flows, which they deem ``elastic'' because they try to grab as much bandwidth as available, from other flows that send at a constant rate, which they call ``unelastic''. They use this elasticity detection to determine if delay-based \gls{cc} can be used or not. If cross traffic is deemed elastic then they use classic loss-based \gls{cc} and otherwise delay-based \gls{cc}.

\subsubsection{Flow startup techniques} Besides the classical TCP slow start algorithm \citep{stevens_tcp_1997} some other proposals were made such as Quick Start \citep{jain_quick-start_2019}, which aims to speed up flow start by routers informing the end point, which rate they deem appropriate. \citep{jarvinen_congestion_2019} investigate the impact of \gls{aqm} and \gls{cc} during flow startup. They inspect a variety of different \gls{aqm} algorithms and also investigate \gls{fq}. \citep{kuhlewind_chirping_2010} propose ``chirping'' for flow startup to estimate the available bandwidth. This works by sending packets with different gaps between them to verify what the actual link speed is. \citep{mittal_recursively_2014} propose speeding up flow startup by using packets of different priority. The link is flooded with low-priority packets and in case the link is not fully utilized during startup, these low-priority packets are going to be transmitted, allowing the starting flow to achieve high throughput. If the link is already saturated, the low-priority packets are going to be discarded and no other flow suffers.

%JF: Evtl.statt subsubsection Absätze mit Titel fett verwenden um doch einigen Platz zu sparen?
\subsubsection{Detecting \gls{aqm} mechanisms} There are few publications that investigate methods to detect the presence of an \gls{aqm}. \citep{kargar_bideh_tada_2016} propose a method to detect and distinguish certain popular \gls{aqm} mechanisms. However, unlike my work, it doesn't consider \gls{fq}. \citep{baykal_detection_2017} propose a machine learning based approach to fingerprint \gls{aqm} algorithm. \gls{fq} is not considered, too.

\section{Concept}
\label{sec:concept_fq}

@@

\begin{figure}
\centering
\subfloat[Sending rate]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_throughput"}.pdf}
\label{fig:throughput}}
\subfloat[Receiving rate (no \gls{fq}).]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_goodput_no_fq"}.pdf}
\label{fig:goodput_no_fq}}
\subfloat[Receiving rate (\gls{fq}).]{\includegraphics[width=0.33\columnwidth]{{"figures/fq_illustration_goodput_fq"}.pdf}
\label{fig:goodput_fq}}
\caption{An example illustrating my proposed flow startup mechanism. Figure \ref{fig:throughput} shows the sending rate, \ref{fig:goodput_no_fq} the receiving rate in case there's no \gls{fq} and \ref{fig:goodput_fq} the receiving rate if there is \gls{fq}.}
\label{fig:illustration}
\end{figure}

The algorithm to determine the presence of \gls{fq} works as follows (seen from the sender's point of view):
\begin{enumerate}
\item Launch two concurrent flows, flow 1 and flow 2 from the same source host to the same destination host. Initialize flow 1 with some starting sending rate and flow 2 with two times that rate. 
\item After every \gls{rtt}, if no packet loss occurred, double the sending rate of both flows. 
\item If packet loss occurred in both flows in the previous \gls{rtt}, calculate the following metric: 
\begin{align}
\textit{loss ratio} &= \frac{\left(\frac{\textit{receiving rate of flow 1}}{\textit{sending rate of flow 1}}\right)}{\left(\frac{\textit{receiving rate of flow 2}}{\textit{sending rate of flow 2}}\right)}\end{align}

This loss ratio metric indicates if flow 2, which has a higher sending rate, achieves a higher receiving rate (goodput) than flow 1. If this is the case, there's no \gls{fq}, otherwise there is.

It is necessary to wait for packet loss in both connections because only then it is certain that the link is saturated. Our algorithm assumes that the link is saturated. 
\item If \gls{fq} is used, the $\textit{loss ratio}$ is 2, otherwise it is 1. I choose 1.5 as the cutoff value. 
\item Since the presence or absence of \gls{fq} is now known the appropriate \gls{cca} can be launched. 
\end{enumerate}

Figure~\ref{fig:throughput}, \ref{fig:goodput_no_fq} and \ref{fig:goodput_fq} schematically illustrate this mechanism. I assume a sender being connected to a receiver, with a bottleneck link between them. I also assume that all flows between the source and the destination follow the same route. On this bottleneck link, the queue is either managed by \gls{fq} or by a shared queue (no \gls{fq}).

Figure~\ref{fig:throughput} shows how the sending rate (observed before the bottleneck link) is doubled after each \gls{rtt}. Furthermore, it shows that flow 2 has double the sending rate of flow 1.

@@

If \gls{fq} is detected, a simple delay-based \gls{cc} is used from then on. Our algorithm is based on the two following simple rules:
\begin{enumerate}
\item If \textit{current rtt} $\leq$ \textit{smallest rtt ever measured on the connection} $+$ 5\textit{ms} then increase the sending rate by 1\%. I choose 5\,ms since a delay of 5\,ms is barely noticeable by humans. This design choice was first established in \citep{nichols_controlling_2012}.
\item Otherwise (\textit{current rtt} > \textit{smallest rtt ever measured on the connection} + 5\textit{ms}) decrease the sending rate by 5\%.
\end{enumerate}

%This \gls{cca} is meant as a simple proof-of-concept to demonstrate the efficacy of my \gls{fq} detection mechanism but I do not claim it to be optimal in any way.

@@

If the absence of \gls{fq} is detected, a loss based \gls{cc} is used as a fallback. I use the algorithm \textit{PCC} proposed by \citep{dong_pcc_2015}. In simplified terms, this \gls{cca} aims to maximize a utility function, which strives for high throughput and low packet loss. However, it is reasonably aggressive and does not starve when competing with other loss-based \glspl{cca} such as Cubic \citep{ha_cubic_2008}.

@@

I base my implementation on the code of \textit{PCC Vivace} \citep{dong_pcc_2018}, which in turn is an improved version of PCC. The PCC code is based on the \textit{UDT} library \citep{gu_udt_2007}, which is a library that provides a reliable transport protocol similar to TCP on top of UDP. This approach is similar to QUIC \citep{iyengar_quic_2018}. An advantage of implementing my approach on top of UDP as opposed to TCP is that it is easier to develop and modify since no kernel module needs to be created and that it is more secure since the code is isolated in a user space process.

@@

Our approach of detecting \gls{fq} relies on using two concurrent flows between the sender and the receiver. During flow startup, user data is transmitted in two flows that the receiver must detect and reassemble. This may be a challenge for deployments since solutions that require cooperation of several instances are more difficult to deploy at a wide scale in the Internet. 
However, multipath TCP \citep{handley_tcp_2012} and QUIC, for which there are proposals to enable recombining data of several flows \citep{de_coninck_multipath_2017}, are becoming increasingly prevalent. I thus argue that the actual obstacles to deployment of my solution are constantly decreasing since QUIC (mostly Google traffic) already accounts for around 
$10\%$ of all Internet traffic \citep{ruth_first_2018} and is growing. Moreover, Apple's macOS and iOS support multipath TCP \citep{apple_about_2018} support.

\section{Experiment setup}

For the evaluation I use the network emulation library \texttt{py-virtnet}\footnote{\url{https://pypi.org/project/py-virtnet/}}. All experiments run on virtual hosts on one physical host, on which the necessary network components and virtual hosts are emulated using Linux network namespaces. The setup consists of two virtual hosts connected through a virtual switch. One virtual host acts as a server and one as a client and a bulk transfer is initiated from the client to the server.

For the experiments with \gls{fq} I use the \texttt{fq} module for Linux. This module keeps one drop tail queue for each flow with a configurable maximum size of packets. However, the design of my \gls{fq} detection algorithm does not assume any particular queue management algorithm being used for the queue of each flow and thus is also compatible with more sophisticated \gls{aqm} mechanisms which were explicitly developed to be used in conjunction with \gls{fq} \citep{taht_flow_2018,hoiland-jorgensen_piece_2018,bachl_cocoa_2019,bachl_lfq_2020}. It only assumes that each flow gets the same share of bandwidth and that for a flow, whose queue is growing faster because more packets of this flow arrive at the bottleneck, packets are dropped faster than for a flow of which fewer packets appear at the bottleneck link. For the experiments with a shared queue I use \texttt{pfifo}, which is a simple drop tail queue which can contain up to a fixed number of packets.

@@

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_just_one_flow_vegas_udp_port9000_fq_10_10_20_1600183241479.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_just_one_flow_vegas_udp_port9000_fq_10_10_20_1600183241479.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by my delay-based \gls{cca} on a link with a speed of 10\,Mbps, a delay of 10\,ms and a buffer size of 100 packets.}
\label{fig:just_one_flow_udp}
\end{figure}

First, I demonstrate empirically that delay-based \gls{cc} actually has a benefit over loss-based one. On one side, \autoref{fig:just_one_flow_tcp} shows that the loss-based \gls{cc} achieves a throughput of almost 10\,Mbps on a 10\,Mbps link, which means full utilization and an \gls{rtt} that is  higher than 100\,ms, which is not desirable. On the other side, \autoref{fig:just_one_flow_udp} shows an example flow using my simple delay-based \gls{cca} and a slow start algorithm similar to the one used by the classic TCP \gls{cca} \textit{New Reno} \citep{allman_tcp_1999}. While the delay-based flow also achieves very high utilization, delay rarely exceeds 20\,ms, meaning that it is more than 5 times lower than the one of the loss-based \gls{cca} of \autoref{fig:just_one_flow_tcp}. The initial peak in \autoref{fig:just_one_flow_udp} is caused by the exponential increase of the sending rate in the beginning, which is commonly used by transport protocols and called ``slow start'' \citep{stevens_tcp_1997}. Our delay-based \gls{cc} also uses a mechanism similar to slow start in the beginning.

\subsection{Comparing flows with different Congestion Controls and AQM mechanisms}
In this scenario, I show an example of a flow with my mechanism that shares a bottleneck link with a flow using a loss-based congestion control.

\subsubsection{No \gls{fq}}
The examples in \autoref{fig:pfifo_tcp} and in \autoref{fig:pfifo_udp} show the loss-based \gls{cca} Cubic compete against my delay-based \gls{cca} on a link without \gls{fq}. It is important to note that I do not use my \gls{fq} detection mechanism here but instead choose delay-based \gls{cc} from the beginning to demonstrate the starvation of delay-based flows that happens when they compete against loss-based ones without \gls{fq}. I start the loss-based flow before to make sure that it has sufficient time to fill up the link: The loss-based flow starts 5 seconds earlier and after 5 seconds, the delay-based flow joins. I choose these 5 seconds so that the first flow can enter its stable state because I want to show how how the flow, even though it had the advantage of the 5 seconds in which it could use the link all for itself, loses to the loss-based \gls{cc}: While the delay-based flow gains some share of the link during startup, it is pushed away by the loss-based flow later on and ``starves''.

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_competing_flow_pfifo_vegas_udp_port9000_pfifo_10_50_20_1600182743408.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_competing_flow_pfifo_vegas_udp_port9000_pfifo_10_50_20_1600182743408.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by my delay-based \gls{cca} on a link with a speed of 50\,Mbps, a delay of 10\,ms and a buffer size of 100 packets. The bottleneck is controlled by a shared queue (no \gls{fq}) and is shared with the flow of \autoref{fig:pfifo_tcp}.}
\label{fig:pfifo_udp}
\end{figure}

@@

\begin{figure}[h]
\centering
\subfloat[Throughput (receiving rate)]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/throughput_1_competing_flow_fq_udp_port9000_fq_10_50_20_1600182657352.pcap"}.pdf}%
}\hfill
\subfloat[\gls{rtt}]{%
  \includegraphics[width=0.48\columnwidth]{{"figures/plots/rtt_1_competing_flow_fq_udp_port9000_fq_10_50_20_1600182657352.pcap"}.pdf}%
}
\caption{Throughput and delay of a flow controlled by my delay-based \gls{cca} on a link with a speed of 50\,Mbps, a delay of 10\,ms and a buffer size of 100 packets. The bottleneck is controlled by \gls{fq} and is shared with the flow of \autoref{fig:fq_tcp}.}
\label{fig:fq_udp}
\end{figure}
